---
title: "Project1.qmd"
format: html
editor: visual
---

# Introduction

This project uses data from the Public use Microdata Sample(PUMS) Census API. The goal is to create functions that can retrieve, process, summarize, and visualize census data from multiple years.

# Data Processing

## Parsing Variable Values

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
census_key <- Sys.getenv("CENSUS_API_KEY")
```

```{r}
get_variable_info <- function(year){
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json")
  response <- GET(url)
  json_text <- content(response, as = "text", encoding="UTF-8")
  variables <- fromJSON(json_text, simplifyVector = FALSE)$variables
  
  if (year %in% c(2021, 2022)){
    state_var <- "ST"
  } else {
    state_var <- "STATE"
  }
  
  wanted_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "HHL", "SCH", "SCHL", "SEX", state_var, "REGION", "DIVISION")
  
  variables[wanted_vars]
}
```

```{r}
variables_2021 <- get_variable_info(2021)
variables_2022 <- get_variable_info(2022)
variables_2023 <- get_variable_info(2023)
variables_2024 <- get_variable_info(2024)

variable_info <- list(
  "2021" = variables_2021,
  "2022" = variables_2022,
  "2023" = variables_2023,
  "2024" = variables_2024
)

names(variable_info)

dir.create("data", showWarnings = FALSE)
saveRDS(variable_info, "data/variable_info.rds")
```

## Parsing API Data

Now that the variable information has been saved, the next step is to query and process the PUMS API data. I created a helper function to convert the API response into a tibble before building the main function used to retrieve Census data.

```{r}
response_to_tibble <- function(response) {
  data <- content(response)
  column_names <- unlist(data[[1]])
  data_rows <- data[-1]
  data_matrix <- do.call(rbind, lapply(data_rows, unlist))
  result <- as_tibble(data_matrix)
  names(result) <- column_names
  return(result)
}
```

The API response stores the column names in the first element and the observations in the remaining elements. I created a helper function to separate these pieces and convert the returned data into a tibble.

```{r}
convert_time <- function(x, value_info) {
  labels <- vapply(x, function(code) {
    if (is.na(code) || code == "0") {
      return(NA_character_)
    }
    
    code <- sprintf("%03d", as.numeric(code))
    label <- value_info[[code]]
    
    if (is.null(label)) {
      return(NA_character_)
    }
    
    return(label)
  }, character(1))
  
  labels <- gsub("a\\.m\\.", "AM", labels)
  labels <- gsub("p\\.m\\.", "PM", labels)
  
  start_time <- sub(" to .*", "", labels)
  end_time <- sub(".* to ", "", labels)
  
  start_time <- lubridate::parse_date_time(
    start_time,
    orders = "I:M p",
    quiet = TRUE
  )
  
  end_time <- lubridate::parse_date_time(
    end_time,
    orders = "I:M p",
    quiet = TRUE
  )
  
  midpoint <- start_time + (end_time - start_time) / 2
  
  return(midpoint)
}
```

```{r}
get_census_data <- function(year = 2024,
                            num_vars = c("AGEP", "PWGTP"),
                            cat_vars = "SEX",
                            geography = "State",
                            geo_value = NULL) {
  
  valid_years <- c(2021, 2022, 2023, 2024)
  
  if (length(year) != 1 || !year %in% valid_years) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  valid_num_vars <- c(
    "AGEP", "GASP", "GRPIP", "JWAP",
    "JWDP", "JWMNP", "PWGTP"
  )
  
  if (!all(num_vars %in% valid_num_vars)) {
    stop("Invalid numeric variable.")
  }
  
  if (!"PWGTP" %in% num_vars) {
    stop("PWGTP must be included.")
  }
  
  if (length(setdiff(num_vars, "PWGTP")) < 1) {
    stop("At least one numeric variable other than PWGTP must be included.")
  }
  
  valid_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  if (length(cat_vars) < 1) {
    stop("At least one categorical variable must be included.")
  }
  
  if (!all(cat_vars %in% valid_cat_vars)) {
    stop("Invalid categorical variable.")
  }
  
  valid_geographies <- c("Region", "Division", "State")
  
  if (length(geography) != 1 || !geography %in% valid_geographies) {
    stop("Geography must be Region, Division, or State.")
  }
  
  if (is.null(geo_value)) {
    if (geography == "State") {
      geo_value <- "37"
    } else if (geography == "Region") {
      geo_value <- "3"
    } else {
      geo_value <- "5"
    }
  }
  
  variable_info <- readRDS("data/variable_info.rds")
  year_info <- variable_info[[as.character(year)]]
  
  if (geography == "State") {
    if (year %in% c(2021, 2022)) {
      geo_var <- "ST"
    } else {
      geo_var <- "STATE"
    }
  } else {
    geo_var <- toupper(geography)
  }
  
  valid_geo_values <- names(year_info[[geo_var]]$values$item)
  
  if (tolower(as.character(geo_value)) != "all" &&
      !as.character(geo_value) %in% valid_geo_values) {
    stop("Invalid geography value.")
  }
  
  if (tolower(as.character(geo_value)) == "all") {
    geo_value <- "*"
  }
  
  requested_vars <- unique(c(num_vars, cat_vars))
  get_vars <- paste(requested_vars, collapse = ",")
  
  census_key <- Sys.getenv("CENSUS_API_KEY")
  
  if (census_key == "") {
    stop("Census API key was not found.")
  }
  
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    get_vars,
    "&for=",
    tolower(geography),
    ":",
    geo_value,
    "&key=",
    census_key
  )
  
  response <- GET(url)
  stop_for_status(response)
  
  result <- response_to_tibble(response)
  
  for (var in setdiff(num_vars, c("JWAP", "JWDP"))) {
    result[[var]] <- as.numeric(result[[var]])
  }
  
  time_vars <- intersect(num_vars, c("JWAP", "JWDP"))
  
  for (var in time_vars) {
    result[[var]] <- convert_time(
      result[[var]],
      year_info[[var]]$values$item
    )
  }
  
  for (var in cat_vars) {
    value_info <- year_info[[var]]$values$item
    
    if (!is.null(value_info)) {
      result[[var]] <- factor(
        result[[var]],
        levels = names(value_info),
        labels = unlist(value_info)
      )
    }
  }
  
  class(result) <- c("census", class(result))
  
  return(result)
}
```

```{r}
test_time <- get_census_data(
  num_vars = c("AGEP", "JWAP", "PWGTP")
)

test_time |>
  filter(!is.na(JWAP)) |>
  select(AGEP, JWAP, PWGTP, SEX) |>
  head()
```

```{r}
test_state <- get_census_data(
  geography = "State",
  geo_value = "37"
)

head(test_state)
```

```{r}
test_all <- get_census_data(
  geography = "State",
  geo_value = "all"
)

head(test_all)
```
